Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-20 manuscript/paper.tex
summary. almost done

\com{
%---------------------------------------------------------
% REVIEWS:
WangS03, %REVIEW
%{Epitaxial BiFeO<sub>3</sub> Multiferroic Thin Film Heterostructures}
DagottoS05,
%{Complexity in Strongly Correlated Electronic Systems},
SchlomARMR07,
%Strain Tuning of Ferroelectric Thin Films}
KhomskiiPhys09
%Classifying multiferroics: Mechanisms and effects}
ZubkoARCMP11,
%{Interface Physics in Complex Oxide Heterostructures},
HwangNM12, 
%{Emergent phenomena at oxide interfaces},

BhattacharyaARMR14,
%{Magnetic Oxide Heterostructures},
HellmanRMP17,
%{Interface-induced phenomena in magnetism},
  ChenJPCM17
 % {Charge transfer driven emergent phenomena in oxide heterostructures},
DagottoMRSB08,
  %{Strongly Correlated Electronic Materials: Present and Future},

SalamonRMP01, 
%The physics of manganites: Structure and transport},
DawberRMP05,
% Physics of thin-film ferroelectric oxides},
KosterRMP12,
%Structure, physical properties, and applications of ${\mathrm{SrRuO}}_{3}$ thin films},

%electronic correlations:
~\cite{DagottoS05,DagottoMRSB08}

%applications:
~\cite{SalamonRMP01,
WangS03
DawberRMP05,
SchlomARMR07,
KosterRMP12}

% understanding:
~\cite{ZubkoARCMP11,
HwangNM12, 
BhattacharyaARMR14,
HellmanRMP17,
ChenJPCM17}

%---------------------------------------------------------

LiN19,
% Superconductivity in an infinite-layer nickelate},
LeeN10,
%{A strong ferroelectric ferromagnet created by means of spin--lattice coupling},
BousquetN08,
%{Improper ferroelectricity in perovskite oxide artificial superlattices},

OjaPRL12,
%{${d}^{0}$ Ferromagnetic Interface between Nonmagnetic Perovskites},

WangPRX16,
% {Stabilization of Highly Polar ${\mathrm{BiFeO}}_{3}$-like Structure: A New Interface Design Route for Enhanced Ferroelectricity in Artificial Perovskite Superlattices}
CavigliaPRL12,
%{Ultrafast Strain Engineering in Complex Oxide Heterostructures},
XiaPRB09,
%{Critical thickness for itinerant ferromagnetism in ultrathin films of ${\text{SrRuO}}_{3}$},

BenedekPRL11,
%{Hybrid Improper Ferroelectricity: A Mechanism for Controllable Polarization-Magnetization Coupling},
GarrityPRL14,
% {Hyperferroelectrics: Proper Ferroelectrics with Persistent Polarization},
KobayashiPRL12,
%{Electronic Ferroelectricity in a Molecular Crystal with Large Polarization Directing Antiparallel to Ionic Displacement},

HarrisPRB07,
% {Landau analysis of the symmetry of the magnetic structure and magnetoelectric interaction in multiferroics},
LeePRL10,
%{Epitaxial-Strain-Induced Multiferroicity in ${\mathrm{SrMnO}}_{3}$ from First Principles},
WangPRL15,
%{Observation of Magnetoelectric Multiferroicity in a Cubic Perovskite System: ${\mathrm{LaMn}}_{3}{\mathrm{Cr}}_{4}{\mathrm{O}}_{12}$},

ChenPRL22,
%{Optically Controlled Femtosecond Polariton Switch at Room Temperature}
ChoiPRL10,
% {Cross-Control of Magnetization and Polarization by Electric and Magnetic Fields with Competing Multiferroic and Weak-Ferromagnetic Phases},
KidaPRB08,
%{Electrically driven spin excitation in the ferroelectric magnet ${\text{DyMnO}}_{3}$},
ChoPRL16,
%{Laser Particle Stimulated Emission Microscopy},
deJongPRL12,
%{Coherent Control of the Route of an Ultrafast Magnetic Phase Transition via Low-Amplitude Spin Precession},

ShirakoPRB11,
%{Integer spin-chain antiferromagnetism of the 4$d$ oxide CaRuO${}_{3}$ with post-perovskite structure},
DikinPRL11,
% {Coexistence of Superconductivity and Ferromagnetism in Two Dimensions},
ChoudhuryPRL12,
%{Near-Room-Temperature Colossal Magnetodielectricity and Multiglass Properties in Partially Disordered ${\mathrm{La}}_{2}{\mathrm{NiMnO}}_{6}$},

LeePRL11,
%  {Electrolyte Gate-Controlled Kondo Effect in ${\mathrm{SrTiO}}_{3}$},

@@

\com{---a cooperative Jahn-Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap}
whereby the orbital degeneracy of the B ions
is lifted that stabilises the distortion---
}

@@

\com{
We take a completely different approach to this problem.
We go back to the basics
and explore the octahedral rotations from the perspective
of (first order) JT effect itself
}

@@

The Jahn--Teller effect~\cite{JahnPRSL37,JahnPRSL38,OpikRS57,Sturge68,GehringRPP75,KugelSPU82,PavariniChap} describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion.

@@

Does this picture survive when we consider the potential of the full ionic lattice at the site of B atom?
Yes.... qualitatively, the same behaviour. just quantitative differences, the balance between A/B cages shifts towards lower $q_A$...

@@

>>>>

@@

@@

@@

The Jahn--Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap,
KhomskiiCR21}
 describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion.

@@

@@

Does this picture survive when we consider the potential of the full ionic lattice at the site of B atom?
Yes.... qualitatively, the same behaviour. just quantitative differences, the balance between A/B cages shifts towards lower $q_A$... 
}

\rev{
consider a few diff $d^{n}$ cases:
1 e: for a given qa,qb; there might be incentives to prefer one angle over the other....
but more interesting cases:
where one angle does not lower the energy of many-electron ground state
while the other does. so, obviously, such systems would take the energetically favorable path...
}

\section{Summary and Conclusion}
We consider the effects of octahedral rotations and tilting in perovskites
on the $t_{2g}$ and $e_g$ manifolds of the central transition metal ions
and find significant splitting in these levels
caused by the distortions in the crystal field of nearest A and B ion cages,
which
essentially amounts to a 
Jahn-Teller instability.
A competition between the two ion cages
determines the sign and strength of this effect
depending on the proportion of charges $q_A,q_B$ on A and B ions.
The tilt 
%$\alpha$ 
and rotation 
%$\beta$
also tend to establish different ground states
that depends on which ion cage dominates.
If A cage dominates (i.e., $q_A$ is larger than a critical value),
%$\alpha$
tilting lowers
$xy,x^2-y^2$ orbitals
and raises $yz,zx,z^2$,
while 
%$\beta$
rotation tends to do the opposite;
the whole situation reverses if B cage dominates.

In a given material,
assuming
the charges on A and B are fixed
and either A or B dominating,
one type of distortion 
would be energetically preferred
over the other
for certain
numbers of electrons in 
the $d$-orbitals of the Jahn-Teller ion B.
For example,
assuming B cage dominating,
a $d^2$ configuration
should prefer tilting over rotation
because it stabilizes two occupied orbitals 
while rotation stabilizes one and destabilizes the other occupied orbital.
